Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2025-01-14 analysis/paper/paper.Rmd
Final edits after peer review

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  # \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/apa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. The precise nature of this discrepancy will necessitate further substantiation and the method of shoreline dating itself still requires further testing and assessment. Having noted this, we still tentatively suggest that while the number of shoreline dated sites is not devoid of influence from a demographic signal, this measure could be more heavily influenced by mobility patterns. Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data more directly reflects population dynamics.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
Keywords: `r rmarkdown::metadata$keywords`

@@

```{r packages, echo = FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(patchwork)
library(kableExtra)
library(sf)
library(shoredate)
```

```{r data, include = FALSE}
surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
sites <- st_read(here("analysis/data/derived_data/combined_sites.gpkg"))
```

# Introduction
Variations in population patterns are regarded as important drivers and potential outcomes of cultural variation, and are thus of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates has become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [e.g. @bergsvik2021; @solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @jorgensen2020b; @lundstrom2023; @persson2018; see also @persson2009].

The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012; @rick1987]. In this paper, we report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the retreat of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or located close to the contemporaneous shoreline at their time of use. Thus, as coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. A series of studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population density [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Fig. \@ref(fig:map)). The result of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

The present study is decidedly exploratory. While some thoughts concerning the relationship between the variables underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies indicate that they do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden for the Museum of Cultural History. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to lose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of 310 ^14^C-dates. Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above.

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.5281/zenodo.10449324. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates was done using the R packages *ADMUR* v1.0.3 [@timpson2021] and *rcarbon* v1.4.4 [@crema2021]. The R package *shoredate* v1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.

@@

```{r map, echo = FALSE, fig.cap = "A) Map displaying the location of the study area and shoreline dated sites. Black lines indicate the borders between municipalities. The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1. B) The distribution of radiocarbon dates employed in the study.", out.width = "500px", fig.align="center", regfloat = TRUE}

@@

To account for investigatory bias that can result from variable sampling intensity between sites, the summing of radiocarbon dates with *ADMUR* starts with binning dates from the same sites that fall within 200 uncalibrated ^14^C years of each other when measured from the mean ^14^C age. All dates are then calibrated using a resolution of 5 years and normalised to sum to unity. The RSPDs for each site-phase are summed to achieve the final SPD, which is in turn normalised. All calibrations were performed using the IntCal20 calibration curve [@reimer2020].

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE estimation with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10,000 times for each null model. The 2.5th and 97.5th percentile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Fig. \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory of relative sea-level (RSL) change has been the same [see @svendsen1987, and map in Fig. \@ref(fig:shoredate)A]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the RSL change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Fig. \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not consider that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Fig. \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore be commutative [see @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

```{r shoredate, echo = FALSE, fig.cap = "A) The location of an example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site. C) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea and represents a \\textit{terminus post quem} date for the site. D) The shoreline dating approach used for this study, where the gamma function on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

The implications of different approaches to shoreline dating and summing the results are illustrated in Fig. \@ref(fig:sumshore). In Fig. \@ref(fig:sumshore)A the sites to be shoreline dated are plotted according to their elevation. Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly inform us about their temporal distribution. For example, as the trajectory for sea-level regression is more pronounced towards the north-east, the same elevation at a location to the south-west implies a younger date compared to one situated further north-east.

When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Fig. \@ref(fig:shoredate)C). As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years. This approach is illustrated in Fig. \@ref(fig:sumshore)B. The SSPD resulting from using the probabilistic method of shoreline dating, given in Fig. \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

```{r sumshore, echo = FALSE, fig.cap = "A) The distribution of the shoreline dated sites (n = 921) by altitude above present sea-level. The binwidth is 4 m. B) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates. C) The result of instead summing the probability of the shoreline dates achieved with the method employed here, which accounts for the likely distance between sites and shoreline. The binwidth is 5 years, following the resolution used with the dating procedure.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probabilities and is not, critically, found by using the values of the final SPD [@crema2022; @timpson2021]. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use to avoid over-fitting, and provides a measure for the relative performance of the models.

@@

# Data, scripts, code, and supplementary information availability{-}
All data, code, text and figures used for this paper is freely available in a version-controlled repository on GitHub at https://github.com/isakro/se.norway.shoredate.14c. This has been archived on Zenodo at https://doi.org/10.5281/zenodo.10449324.

@@

<div id="refs"></div>

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Department of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      extra_dependencies: ["float"]
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  # \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/apa.csl" # Insert path for the bib-style
abstract: |
  By developing a new methodology for handling and assessing a large number of shoreline dated sites, this paper compares the summed probability distribution of radiocarbon dates and shoreline dates along the Skagerrak coast of south-eastern Norway. Both measures have previously been compared to elucidate demographic developments in Fennoscandia, but these have not been based on probabilistic methods for shoreline dating. The findings indicate a largely diverging development of the two data sets through the Mesolithic. The number of shoreline dated sites undergoes some process of overall decrease through the period, while the radiocarbon data is characterised by a lacking signal in the earliest parts of the period and then undergoes a logistic growth that quickly plateaus and remains stable for the remainder of the period. The precise nature of this discrepancy will necessitate further substantiation and the method of shoreline dating itself still requires further testing and assessment. Having noted this, we still tentatively suggest that while the number of shoreline dated sites is not devoid of influence from a demographic signal, this measure could be more heavily influenced by mobility patterns. Conversely, we also suggest that the lacking radiocarbon data from the earliest part of the Mesolithic is in part the result of mobility patterns, but that the radiocarbon data could be reflecting population dynamics more directly.
keywords: | 
  Mesolithic; Shoreline dates; Radiocarbon dates; Fennoscandia
---
#' \pagenumbering{gobble}

Keywords: `r rmarkdown::metadata$keywords`

@@

```{r packages, echo = FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(patchwork)
library(kableExtra)
library(sf)
```

```{r data, include = FALSE}
surveyed <- st_read(here("analysis/data/raw_data/surveyed_sites.gpkg"))
sites <- st_read(here("analysis/data/derived_data/combined_sites.gpkg"))
log_params <- read.csv(here("analysis/data/derived_data/logistic_parameters.csv"))
load(here("analysis/data/derived_data/c14_details.RData"))

```

# Introduction
Population patterns are regarded as important drivers and potential outcomes of cultural variation, and are thus of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates has become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [e.g. @bergsvik2021; @solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @jorgensen2020b; @lundstrom2023; @persson2018; see also @persson2009].

The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012; @rick1987]. In this paper, we report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). While fundamental questions concerning the reliability of shoreline dating still persist, our analysis draws on recent developments and ongoing efforts to formalise and better understand the dating method [@roalkvam2023; @rosenvinge2022, 21--41]. We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the retreat of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerrak coast are interpreted as having been shore bound or located close to the contemporaneous shoreline at their time of use [see e.g. @darmark2018a; @solheim2020a; @mjærum2022; @roalkvam2023]. Consequently, as coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the occupation of the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Fennoscandian archaeology [@brogger1905; @hollender1901], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. A series of studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population density [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. These point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Fig. \@ref(fig:map)). The result of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Consequently, these geographical and temporal limits are also used for this study.

Limiting this study to the same temporal and geographical constraints as was used in @roalkvam2023 should mitigate some issues that could follow from undue extrapolation of the technique to areas where the reliability of the method and the parametrisation employed here has not been assessed and therefore might not apply. However, while we view the long history of use and research into shoreline dating as a testament to its general applicability, and these recent developments as an improvement of this framework, it should also be underscored that the method for shoreline dating employed in this study is still hefted with significant questions regarding its reliability [see @roalkvam2023].

The present study is decidedly exploratory. While some thoughts concerning the relationship between the frequency of radiocarbon dates, shoreline dated sites and population dynamics underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between these variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies indicate that they do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and an unpublished data set collated by Solheim and Kjetil Loftsgarden for the Museum of Cultural History. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to lose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to a final sample of `r all_included` ^14^C-dates. Probabilities falling after 4000 BCE, and the first possible small-scale introduction of agriculture in south-eastern Norway, were retained to account for edge-effects but these results should be treated with care for the reasons noted above. Of the final sample, `r char_nut` (`r round((char_nut/all_included)*100)`%) dates are made on charcoal and burnt nutshell and `r other` (`r round((other/all_included)*100)`%) on other material such as bone.

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in an online repository at https://doi.org/10.5281/zenodo.10449324. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates was mainly done using the R package *ADMUR* v1.0.3 [@admur2020; @timpson2021] and with some use of *rcarbon* v1.4.4 [@crema2021]. The R package *shoredate* v1.0.2 was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* used for subsequent model fitting, Monte Carlo simulation and model comparison.

@@

```{r map, echo = FALSE, fig.cap = "a) Map displaying the location of the study area and shoreline dated sites. Black lines indicate the borders between municipalities. The surveyed sites included in the analysis are the ones given a quality score of 3 or higher using the framework in Table 1. b) The distribution of radiocarbon dates employed in the study.", out.width = "500px", fig.align="center", regfloat = TRUE}

@@

To account for investigatory bias that can result from variable sampling intensity between sites, the summing of radiocarbon dates with *ADMUR* starts with binning dates from the same sites that fall within 200 uncalibrated ^14^C years of each other when measured from the mean ^14^C age. All dates are then calibrated using a resolution of 5 years and normalised to sum to unity. The RSPDs for each site-phase are summed to achieve the final SPD, which is in turn normalised. All calibrations were performed using the IntCal20 calibration curve [@reimer2020]. It can be noted that while several authors have pointed to the fact that normalising the dates can lead to spikes in the RSPD [e.g. @crema2021, 27--29; @weninger2015], not normalising the dates did not result in any meaningful difference in the resulting best-fit models for the purposes of this paper (see supplementary figure).

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE estimation with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of site phases in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10,000 times for each null model. The 2.5th and 97.5th percentile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Fig. \@ref(fig:shoredate)b). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory of relative sea-level (RSL) change has been the same [see @svendsen1987, and map in Fig. \@ref(fig:shoredate)a]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the RSL change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Fig. \@ref(fig:shoredate)c indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the common practice of adding a constant error or binning point estimates of shoreline dates, as mentioned above, does not consider that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Fig. \@ref(fig:shoredate)d). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore be commutative [see @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

```{r shoredate, echo = FALSE, fig.cap = "a) The location of an example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  b) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site. c) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea and represents a \\textit{terminus post quem} date for the site. d) The shoreline dating approach used for this study, where the gamma function on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

The implications of different approaches to shoreline dating and summing the results are illustrated in Fig. \@ref(fig:sumshore). In Fig. \@ref(fig:sumshore)a the sites to be shoreline dated are plotted according to their elevation. Due to the variable rates of shoreline displacement within the study region, the same elevation does not necessarily equate to the same date, and so this does not directly inform us about their temporal distribution. For example, as the trajectory for sea-level regression is more pronounced towards the north-east, the same elevation at a location to the south-west implies a younger date compared to one situated further north-east.

When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point estimates of dates by either using the mean of the displacement curves, or by taking the mean of the TPQ dates for each site (as illustrated in Fig. \@ref(fig:shoredate)c). As outlined above,  these point estimates are then typically aggregated in bins of 200 or 500 years. This approach is illustrated in Fig. \@ref(fig:sumshore)b. The SSPD resulting from using the probabilistic method of shoreline dating, given in Fig. \@ref(fig:sumshore)c, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

```{r sumshore, echo = FALSE, fig.cap = "a) The distribution of the shoreline dated sites (n = 921) by altitude above present sea-level. The binwidth is 4 m. b) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates. c) The result of instead summing the probability of the shoreline dates achieved with the method employed here, which accounts for the likely distance between sites and shoreline. The binwidth is 5 years, following the resolution used with the dating procedure.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probabilities and is not, critically, found by using the values of the final SPD [@crema2022; @timpson2021]. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use to avoid over-fitting and provides a measure for the relative performance of the models.

@@

# Data, scripts, code, and supplementary information availability{-}
All data, code, text and figures used for this paper are freely available in a version-controlled repository on GitHub at https://github.com/isakro/se.norway.shoredate.14c. This has been archived on Zenodo at https://doi.org/10.5281/zenodo.8373856

@@

<div id="refs"></div>
\newpage
